\thesistitle{A High and Variable Dimensional Measurement of the $Z$+jets Differential Cross Section with the ATLAS Experiment and Aritifical Intelligence}

%"Dissertation" for PhD, "Thesis" for master's
\documenttitle{Dissertation}

\degreename{Doctor of Philosophy}

% Use the wording given in the official list of degrees awarded by UCI:
% http://www.rgs.uci.edu/grad/academic/degrees_offered.htm
\degreefield{Physics and Astronomy}

% Your name as it appears on official UCI records.
\authorname{Kevin Greif}

% Use the full name of each committee member and full title 
% (e.g. Professor/Associate Professor).
\committeechair{Daniel Whiteson}
\othercommitteemembers
{
  Andrew Lankford \\
  Jonathan Feng
}

\degreeyear{2026}

\copyrightdeclaration
{
  {\copyright} {\Degreeyear} \Authorname
}

% If you have previously published parts of your manuscript, you must list the
% copyright holders; see Section 3.2 of the UCI Thesis and Dissertation Manual.
% Otherwise, this section may be omitted.
% \prepublishedcopyrightdeclaration
% {
% 	Chapter 4 {\copyright} 2003 Springer-Verlag \\
% 	Portion of Chapter 5 {\copyright} 1999 John Wiley \& Sons, Inc. \\
% 	All other materials {\copyright} {\Degreeyear} \Authorname
% }

% The dedication page is optional
% (comment out to exclude).
\dedications
{
  (Optional dedication page)
  
  To Dario Amodei mostly I guess?
}

\acknowledgments
{
  I would like to thank...
  
  (You must acknowledge grants and other funding assistance. 
  
  You may also acknowledge the contributions of professors and
  friends.
  
  You also need to acknowledge any publishers of your previous
  work who have given you permission to incorporate that work
  into your dissertation. See Section 3.2 of the UCI Thesis and
  Dissertation Manual.)
}


% Curriculum vitae is in vita.tex, which is \input into main.tex.

% The abstract was previously limited to a maximum of 350 words, 
% but the UCI manual at https://etd.lib.uci.edu/electronic/td2e#2.2.1.
% currently does not indicate that there is any word limit for the abstract
\thesisabstract
{
  Proton-proton collisions at the Large Hadron Collider (LHC) offer the opportunity to observe the interactions of fundamental particles at very high energy scales. The high collision rate and large number of final state particles produced per collision imply that the datasets produced by detectors such as the ATLAS experiment are large and highly dimensional. The complexity of these datasets calls for the use of novel data analysis techniques which can exploit all of the available information to illuminate known interactions and search for new ones. This thesis presents applications of artifical intelligence (AI) techniques to improve the physics results of the ATLAS experiment. It centers on a full phase measurement of the Z+jets production cross section at the LHC, where the cross section is measured differential in the kinematics of every final state charged particle through the use of an AI based unfolding algorithm. This is the first such measurement performed at the LHC, which provides a complete experimental characterisation of the Z+jets production process and a proof-of-principle for the further pursuit of such measurements on a host of LHC processes.
}


%%% Local Variables: ***
%%% mode: latex ***
%%% TeX-master: "thesis.tex" ***
%%% End: ***
