% ============================================================
% vita.tex -- Curriculum Vitae for the UCI thesis
%
% The \curriculumvitae{} macro (defined in ucithesis.cls) stores
% its argument and renders it on the VITA preliminary page.
% This file is \input into main.tex before \begin{document}.
%
% UCI requirement: include at minimum education (with dates and
% institutions) plus the degree currently being attained.
% ============================================================

% ----------------------------------------------------------
% Helper commands used only in the vita
% ----------------------------------------------------------

% \vitaentry{Title}{Date(s)}{Description}
% Produces a two-column entry: bold title flush-left, bold date flush-right,
% followed by an italic description.
\newcommand{\vitaentry}[3]{%
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lr}
    \textbf{#1} & \textbf{#2}
  \end{tabular*}\\[-0.3em]
  \textit{#3}\vspace{0.8em}
}

% \pubentry{Authors}{Title}{Venue}{Year}
% Single numbered publication entry.
\newcommand{\pubentry}[4]{\item #1. ``#2''. #3, #4.}

% ----------------------------------------------------------
% Vita content
% ----------------------------------------------------------

\curriculumvitae{

% --- Education ---
\textbf{EDUCATION}
\vspace{0.5em}

\vitaentry{Doctor of Philosophy in Physics and Astronomy}{2026}%
  {University of California, Irvine \hfill Irvine, California \\
   Advisor: Daniel Whiteson}

\vitaentry{Bachelor of Science, Magna Cum Laude}{2020}%
  {University of Notre Dame \hfill Notre Dame, Indiana \\
   Majors: Physics and Program of Liberal Studies}

\vspace{0.5em}

% --- Research Experience ---
\textbf{RESEARCH EXPERIENCE}
\vspace{0.5em}

\vitaentry{Foundation models for particle physics}{Ongoing}%
  {Researched the application of pre-trained foundation models in particle physics. Compared their performance to equivariant neural networks, and investigated the proper composition of pre-training data to optimize scaling laws.}

\vitaentry{Artificial Intelligence based unfolding}{Ongoing}%
  {Developed novel unfolding methods based on generative models
   and contributed to community comparison studies. Studied the use of
   unbinned, highly differential cross section measurements for parameter
   estimation. Deployed Artificial Intelligence based unfolding to perform a
   hyper-differential cross section measurement of $Z$+jets production
   with the ATLAS detector.}

\vitaentry{Constituent based top tagging}{2022--2024}%
  {Benchmarked several constituent top-tagging algorithms on realistic detector
   simulation. Assessed systematic uncertainties in tagger performance.
   Prepared research-grade open data records for use by the broader particle
   physics community.}

\vitaentry{Calibration of the ATLAS New Small Wheel}{2022--2023}%
  {Calibrated the peak detector output (PDO) readout used in the ATLAS New Small
   Wheel muon detector. Participated in detector operations during Run 3.}

\vitaentry{Physics inspired neural networks}{2018--2020}%
  {Deployed a physically motivated neural network architecture on a
   classification problem in particle physics.}

\vitaentry{CMS track trigger development}{2017--2019}%
  {Improved the efficiency of cuts on track candidates in the CMS experiment's
   track trigger algorithm.}

\vspace{0.5em}

% --- Publications ---
\textbf{SELECTED PUBLICATIONS}
\vspace{0.5em}

{\small Only includes papers in which I made a significant contribution,
in reverse chronological order.}
\vspace{0.3em}

\begin{enumerate}
  \pubentry{V.~Breso-Pla et al.}%
    {Explicit or Implicit? Encoding Physics at the Precision Frontier}%
    {\textit{Preprint:} arXiv:2603.08802}{2026}
  \pubentry{A.~Petitjean et al.}%
    {Generative unfolding of jets and their substructure}%
    {\textit{Preprint:} arXiv:2510.19906}{2025}
  \pubentry{A.~Shmakov et al.}%
    {Full event particle-level unfolding with variable-length latent variational diffusion}%
    {\textit{SciPost Phys.}~18(4):117}{2025}
  \pubentry{N.~Huetsch et al.}%
    {The landscape of unfolding with machine learning}%
    {\textit{SciPost Phys.}~18(2):070}{2025}
  \pubentry{ATLAS Collaboration}%
    {Accuracy versus precision in boosted top tagging with the ATLAS detector}%
    {\textit{JINST}~19(08):P08018}{2024}
  \pubentry{A.~Shmakov et al.}%
    {End-to-end latent variational diffusion models for inverse problems in high energy physics}%
    {\textit{NeurIPS Proceedings}}{2023}
  \pubentry{K.~Greif and K.~Lannon}%
    {Physics Inspired Deep Neural Networks for Top Quark Reconstruction}%
    {\textit{EPJ Web Conf.}~245:6029}{2020}
\end{enumerate}

\vspace{0.5em}

% --- Presentations ---
\textbf{PRESENTATIONS}
\vspace{0.5em}

\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}p{2.2cm}p{0.78\textwidth}}
  \textbf{Jul.\ 2025} & ``Forward folding versus unfolding in the age of ML'', \textit{ML4Jets 2025} \\[0.4em]
  \textbf{Nov.\ 2024} & ``Full Event Particle-Level Unfolding with Variable Length Variational Latent Diffusion'', \textit{ML4Jets 2024} \\[0.4em]
  \textbf{Nov.\ 2023} & ``Systematic Effects in Jet Tagging Performance for the ATLAS Detector'', \textit{ML4Jets 2023} \\[0.4em]
  \textbf{Nov.\ 2023} & ``End-to-End Latent Variational Diffusion Models for Unfolding LHC Events'', \textit{ML4Jets 2023} \\[0.4em]
  \textbf{Nov.\ 2022} & ``Constituent-Based Top-Quark Tagging with the ATLAS Detector'', \textit{ML4Jets 2022} \\[0.4em]
  \textbf{Nov.\ 2019} & ``Physics Inspired Deep Neural Networks for Top Quark Reconstruction'', \textit{CHEP 2019} \\[0.4em]
\end{tabular*}

\vspace{0.5em}

% --- Appointments & Memberships ---
\textbf{APPOINTMENTS AND MEMBERSHIPS}
\vspace{0.5em}

\vitaentry{ATLAS Early Career Scientist Board}{2024--Present}%
  {Advocate for the interests of early career scientists within the ATLAS
   collaboration. Organize educational and career development events.}

\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lr}
  \textbf{Phi Beta Kappa Society}  & \textbf{2020--Present} \\
  \textbf{Sigma Pi Sigma Society}  & \textbf{2020--Present} \\
\end{tabular*}

\vspace{0.5em}

% --- Awards ---
\textbf{AWARDS}
\vspace{0.5em}

\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}p{0.55\textwidth}r}
  \textbf{Outstanding Physics Major}  & \textbf{2020} \\
  \multicolumn{2}{p{0.95\textwidth}}{\textit{Physics Department, University of Notre Dame}} \\[0.4em]
  \textbf{Paul Chagnon Award in Physics}  & \textbf{2020} \\
  \multicolumn{2}{p{0.95\textwidth}}{\textit{Physics Department, University of Notre Dame --- recognizing service to the department and fellow students}} \\[0.4em]
  \textbf{Edward J.\ Cronin Award}  & \textbf{2020} \\
  \multicolumn{2}{p{0.95\textwidth}}{\textit{Program of Liberal Studies, University of Notre Dame --- awarded for the finest piece of written work submitted in an academic year}} \\[0.4em]
  \textbf{Hichwa Fellowship}  & \textbf{2019} \\
  \multicolumn{2}{p{0.95\textwidth}}{\textit{Physics Department, University of Notre Dame}} \\
\end{tabular*}

\vspace{0.5em}

% --- Teaching ---
\textbf{TEACHING EXPERIENCE}
\vspace{0.5em}

\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lr}
  \textbf{Teaching Assistant, University of California, Irvine} & \textbf{2020--2021} \\
\end{tabular*}

\begin{itemize}
  \item Physics 3A (Mechanics), Fall 2021
  \item Physics 3LB (Electricity and Magnetism Lab), Spring 2021
  \item Physics 7LC (Mechanics Lab), Winter 2021
  \item Physics 2 (Mathematical Methods in Physics), Fall 2020
  \item Computational Methods in Physics, Spring 2020
\end{itemize}

\vspace{0.5em}

% --- Outreach ---
\textbf{OUTREACH}
\vspace{0.5em}

\vitaentry{ATLAS Virtual Visit Tour Guide}{2022--2025}%
  {Remote visits to the ATLAS experiment for students from around the world.}

\vitaentry{UCI Physics and Astronomy Blog}{2020--2022}%
  {Published accessible descriptions of new research from the UCI physics
   department.}

} % end \curriculumvitae
