
\chapter{Machine Learning and Artificial Intelligence}
\label{ch:ml}

% ---------------------------------------------------------------
% WRITING NOTES:
%   Target: ~15–20 pages (Weeks 3–4: Mar 30 – Apr 12)
%   Goal: give the reader the ML toolkit needed to understand
%         the VLD unfolding architecture (Ch.5) and Omnifold.
%         This is the part of the thesis you most want to
%         remember from your PhD — make it count.
%   Tone: conceptual first, mathematical as needed; always
%         connect back to the HEP application.
% ---------------------------------------------------------------

This Chapter serves as a general introduction to the deep learning methods used throughout the remainder of the thesis\footnote{Note that this can be thought of as the ``instrumentation'' chapter of this thesis, which might otherwise be dedicated to information on the ATLAS detector.}.
Like \Cref{ch:qcd}, the goal is not to be thorough.
Instead only the information needed to understand the particular deep learning methods applied to particle physics data in this thesis will be covered.

\section{Overview: ML as a Paradigm Shift}
\label{sec:ml-overview}

Machine learning is the idea that complex rules or behaviors can be learned directly from data by optimization algorithms, rather than programmed explicitly by human experts.
In classical data analysis, a domain expert encodes their knowledge as a fixed set of rules that are then applied to data to extract a quantitative summary of the information.
It is worth emphasizing that this approach often works very well.
Particle physicists have become very adept at choosing relevant features of the data and learning from them.
However this approach is rarely optimal, especially in cases where the optimal solution exists but is too complex to specify by hand.
To draw an example from \Cref{ch:tagging}, no expert-designed feature set captures the correct decision-surface for classifying jets initiated by boosted top quarks from the light quark and gluon jet background.
This is due to a lack of knowledge about certain features of the jet formation process (e.g. hadronization), but also just due to complexity.
Specifying the optimal feature by hand is probably not possible, but machine learning methods allow this complexity to be learned from the data.

Machine learning methods organize naturally into several paradigms based on the structure of the available training information:

\begin{itemize}
    \item \textit{Supervised learning} trains a model using labeled examples, where each input is paired with a target output.
    Classification and regression are the canonical supervised tasks, and the most widely used in particle physics.
    \Cref{ch:tagging} is entirely dedicated to the jet tagging classification task.
    \item \textit{Unsupervised learning} operates without labels, with the goal of learning structure in the input data. Generative models that learn to produce samples from $p_{\text{data}}$ without access to class labels are a prominent example and are the focus of much of \cref{ch:unfolding}.
    \item \textit{Self-supervised learning} is a hybrid approach in which the supervisory signal is derived from the data itself rather than from external annotation. This is most commonly used to learn general-purpose representations rather than to solve a specific downstream task. In HEP, self-supervised pretraining underlies some recent foundation models for jet physics, which use techniques such as masked particle modeling to learn representations of jet constituents that transfer across tasks~\cite{Golling:2024abg,Birk:2024knn}.
    \item \textit{Reinforcement learning} (RL) trains an agent to maximize a reward signal through interaction with an environment. The key feature of RL is that it does not require a differentiable training objective, though some methods construct a surrogate in practice. RL is responsible for landmark results in game playing~\cite{Silver:2016alphago} and is the basis for the post-training pipelines of modern large language models~\cite{Ouyang:2022instructgpt}. Direct applications of RL within HEP have so far been limited~\cite{Baretz:2025zsv}, though its relevance is growing alongside the agentic AI systems discussed in \cref{ch:conclusion}.
\end{itemize}

Machine learning is not a new idea in particle physics.
Boosted decision trees (BDTs) have been in use in HEP since 2004, and were applied to the single top quark discovery at the Tevatron~\cite{D0:2006ngk,CDF:2009itk}.\footnote{The history of BDTs in HEP is interesting.
The original application of BDTs illustrated that they \textit{outperformed} artificial neural networks.
Following the deep learning revolution initiated by AlexNet, BDTs were largely supplanted by neural networks, which proved far more expressive on the high-dimensional data common in HEP.
Quite recently, BDTs have seen something of a resurgence in recent years, driven by their remarkable training speed relative to neural networks, and their robustness against noisy or irrelevant features in the density ratio estimation tasks covered in \cref{sec:dre}~\cite{Finke:2023ltw}.}
What changed with the deep learning era was not the concept of learned classifiers but the scale of the models.
Neural networks scale better than BDTs, and so could handle the more complicated tasks that have driven much of the interest in ML in recent years.

The success of machine learning in HEP rests on two structural advantages.
The first is access to high-quality simulated data produced by the Monte Carlo event generators discussed in \cref{sec:jets}.
These simulated datasets carry \textit{truth level} information that can be used to derive effective training labels.
This is a significant advantage over other application domains where manual labeling is a significant bottleneck.
Label assignment in HEP has its own subtleties, which are discussed in the context of top tagging in \Cref{ch:tagging}, but effective labels can often be set using straightforward generator-level rules.
The second advantage is that HEP data is generated by physical processes that are well understood and are known to respect specific symmetries.
Lorentz invariance, permutation invariance of jet constituents, and translational equivariance in the $(\eta, \phi)$ plane are all properties that have been used to aid neural network optimization.
These properties could equally well be learned from the data and the community is starting to explore this direction, but exploiting them was a major benefit to early deep learning applications in HEP.

Given that deep neural networks are the dominant paradigm in machine learning at present, the remainder of this chapter provides an overview of the core network architectures and training methods used throughout this thesis.

\section{Neural Networks and Deep Learning}
\label{sec:dl}

% ---------------------------------------------------------------
% Sub-topics to cover:
%
% 1. Multi-layer perceptrons (MLPs)
%    - Composition of affine + nonlinear layers.
%    - Universal approximation theorem (intuition only).
%    - Activation functions: ReLU, GELU.
%
% 2. Training basics
%    - Loss function; stochastic gradient descent; backprop.
%    - Regularization: dropout, weight decay, batch norm.
%    - Practical tips: learning rate scheduling, early stopping.
%
% 3. Convolutional Neural Networks (CNNs)
%    - Translational equivariance; weight sharing.
%    - Early HEP use: jet images (Cogan et al., de Oliveira et al.)
%
% 4. Graph Neural Networks (GNNs)
%    - Jets as point clouds / graphs; node + edge features.
%    - Message passing; DeepSets as a special case (sum aggregation).
%    - ParticleNet (DGCNN-based), PELICAN as examples.
%
% 5. Transformers
%    - Self-attention: queries, keys, values.
%    - Permutation invariance for free — natural for jet constituents.
%    - Particle Transformer (ParT) — state-of-the-art jet tagger.
%    - Connection to large pre-trained models (forward pointer to Ch.4).

\section{Generative Models}
\label{sec:generative}

% ---------------------------------------------------------------
% Sub-topics to cover:
%
% 1. Variational Autoencoders (VAEs)
%    - Encoder q_\phi(z|x), decoder p_\theta(x|z).
%    - ELBO objective; reparameterization trick.
%    - Limitation: blurry samples due to pixel-wise MSE.
%    - HEP use: fast calorimeter simulation (e.g., CaloVAE).
%
% 2. Generative Adversarial Networks (GANs)
%    - Generator vs. discriminator; minimax game.
%    - Training instability; mode collapse.
%    - HEP use: CaloGAN, jet generation.
%
% 3. Normalizing flows
%    - Bijective f: x <-> z with tractable Jacobian.
%    - Change of variables formula: log p(x) = log p(z) + log|det J|.
%    - Coupling layers (RealNVP, Glow); autoregressive flows (MAF).
%    - HEP use: fast simulation (SurVAE, NF-based shower generators),
%      inference (nflows for posterior estimation).
%    - Key property: exact density evaluation — useful for reweighting.
%
% 4. Motivation for moving to diffusion
%    - Flows require bijective transformations → architectural
%      constraints.  GANs are unstable.  VAEs are blurry.
%    - Diffusion models sidestep these issues at the cost of
%      sequential sampling.  Forward pointer to \cref{sec:diffusion}.

\section{Diffusion Models}
\label{sec:diffusion}

% ---------------------------------------------------------------
% Sub-topics to cover:
%
% 1. Score-based / SMLD models
%    - Noise perturbation at multiple scales; score matching.
%    - Song \& Ermon (2019): Generative Modeling by Estimating
%      Gradients of the Data Distribution.
%    - Score function s_\theta(x) ≈ \nabla_x \log p(x).
%
% 2. Denoising Diffusion Probabilistic Models (DDPMs)
%    - Forward process: q(x_t | x_{t-1}) = N(x_t; sqrt(1-beta)x_{t-1}, beta I).
%    - Reverse process: p_\theta(x_{t-1} | x_t) learned by a UNet.
%    - Ho et al. (2020) — key equations; ELBO derivation (high level).
%    - Connection to score matching via x_0-prediction vs. eps-prediction.
%
% 3. Latent Diffusion Models (LDMs)
%    - Two-stage: (a) train a VAE to compress x -> z;
%                 (b) train a diffusion model in latent space z.
%    - Rombach et al. (2022) — Stable Diffusion; key architectural
%      choices: cross-attention for conditioning.
%    - Why latent space? Much lower dimensionality → faster training
%      and sampling; also aligns with the low ID of jet data (Ch.2).
%
% 4. The VLD architecture for unfolding
%    - Forward pointer to Ch.5 where VLD is described in full.
%    - Mention here: VLD = Variational Latent Diffusion, designed
%      specifically for particle physics unfolding.  It conditions
%      the diffusion model on detector-level observables and generates
%      samples from the inferred particle-level distribution.
%    - Highlight what makes VLD novel: unbinned, high-dimensional,
%      generative output that can be histogrammed in any observable.

\section{Density Ratio Estimation and Reweighting}
\label{sec:dre}

% ---------------------------------------------------------------
% Sub-topics to cover:
%
% 1. The core statistical trick
%    - Given two distributions p(x) and q(x), we often want
%      the ratio r(x) = p(x)/q(x) without knowing either
%      density explicitly.
%    - Key insight: a binary classifier trained to distinguish
%      samples from p vs. q learns r(x) implicitly, because
%      the optimal classifier output is r(x)/(1 + r(x)).
%
% 2. Classifier-based density ratio estimation
%    - Train f_\theta(x) -> [0,1] with BCE loss.
%    - Recover r(x) = f(x)/(1 - f(x)).
%    - Works in high dimensions where direct density estimation fails.
%
% 3. Applications in HEP
%    - Monte Carlo reweighting: correct simulation to data.
%    - Importance sampling: reweight one generator to another.
%    - Omnifold (Ch.5): iterative density ratio estimation for
%      unbinned unfolding — this is the key application.
%    - CARL (Cranmer et al.) as a foundational reference.
%
% 4. Practical considerations
%    - Calibration of the classifier output is crucial.
%    - Extreme weights can cause variance blow-up: use weight
%      clipping or regularization.
%    - Choice of architecture matters: must have good coverage of
%      the support of both distributions.

\section{Uncertainty Quantification}
\label{sec:uq}

% ---------------------------------------------------------------
% Sub-topics to cover:
%
% 1. Why UQ matters in HEP
%    - Physics measurements require not just a central value but
%      a rigorous uncertainty.  ML methods add a new source of
%      uncertainty (model uncertainty) on top of statistical and
%      systematic uncertainties.
%    - The ATLAS top tagger paper (Ch.4) deals explicitly with
%      how to assign and propagate ML model uncertainties.
%
% 2. Types of uncertainty
%    - Aleatoric (irreducible, data noise) vs. epistemic (reducible,
%      model / training uncertainty).
%    - In HEP: statistical uncertainty (finite data) is aleatoric;
%      choice of MC generator, network architecture, training
%      procedure are epistemic.
%
% 3. Ensemble methods
%    - Train N models with different seeds / architectures.
%    - Variance of ensemble predictions estimates epistemic uncertainty.
%    - Used in the top tagger paper: ensemble of taggers to
%      derive model uncertainty on the WP efficiency.
%
%
% 5. Systematic uncertainties from ML
%    - Variation of training hyperparameters.
%    - Dependence on MC generator (Pythia vs. Herwig).
%    - "Generator uncertainty" is a major theme in Ch.4 and Ch.6.
