\chapter{Introduction}
\label{ch:intro}

% ---------------------------------------------------------------
% WRITING NOTES:
%   Target: ~4–6 pages (Week 1: Mar 16–22)
%   Arc: open with the two "moments" of 2012, zoom out to the
%        state of HEP open problems, then pivot to ML revolution.
%        End with a roadmap of the thesis.
% ---------------------------------------------------------------

% ---------------------------------------------------------------
% Opening: The two defining moments of 2012
% ---------------------------------------------------------------
%

The discovery of the Higgs Boson, announced on July 4th, 2012~\cite{ATLAS:2012yve,CMS:2012qbp}, is widely regarded as the most important discovery in particle physics in the 21st century.
With that discovery the electroweak symmetry breaking mechanism~\cite{PhysRevLett.13.321, PhysRevLett.13.508, PhysRevLett.13.585} received experimental confirmation and the Standard Model of particle physics~\cite{PhysRevLett.19.1264, THOOFT1972189} was complete.
This triumph of modern science was rightly front-page news, and is still well known in the popular imagination.
A less well known, though equally important, advancement was announced 87 days later on September 30th, 2012 when the results of the ``ImageNet Large Scale Visual Recognition Challenge'' were released.
This was a challenge within the computer vision research community to develop a machine learning model that could sort images into 1000 classes.
The training set for this challenge consisted of 1.2 million painstakingly hand-labeled images, which was unprecedented scale for labeled datasets at the time.
This scale and the recent release of Nvidia's CUDA toolkit, which allowed for optimized matrix operations on GPU hardware, allowed for a team of researchers from the University of Toronto to train a deep convolutional neural network named AlexNet that outperformed its closest competitor by 10\% in top-5 error rate\footnote{The top-5 error rate is the percentage of images that were not correctly classified into the top 5 most likely classes.}~\cite{krizhevsky2012imagenet}.
Before this moment state-of-the-art computer vision models relied on hand-engineered features programmed by human experts, but AlexNet learned all features used to classify the images directly from the data, requiring the humans who programmed it to know essentially nothing about the images themselves.
This breakthrough became known as the ``AlexNet moment'' and marked the beginning of the modern era of machine learning.

This thesis is an attempt to understand the impact of the AlexNet moment on particle physics research in the post-Higgs era.
It's worth dwelling very briefly on the fact that both discoveries lead to a Nobel prize in physics, the Higgs for Peter Higgs and François Englert in 2013 and AlexNet for Geoffrey Hinton in 2024.
The nobel prize for Higgs and Englert was, by my count, the 12th Nobel prize awarded for the development of the Standard Model\footnote{Notable exlusions from this count are the Nobel prizes of Feynman, Schwinger, and Tomonaga for the QED theory which is a precursor of the Standard Model, and the Nobel prizes of Kajita and McDonald for the discovery of neutrino oscillations, which is notably not a part of the Standard Model.}.
It was the capstone of a very fruitful and now very mature research program.
By contrast the Nobel for AlexNet was at least in-part an ``instrumentation'' award\footnote{An excellent parallel to AlexNet is Charles Wilson's development of the cloud chamber, which was a tool developed by a meteorologist that became very useful for fundamental physics research and was awarded the Nobel a decade after it was developed.}, significant for the other discoveries it enables.
Particle physics was one of the first scientific disciplines to make use of deep learning.
The first papers demonstrating the use of deep learning for event selection~\cite{Baldi:2014kfa} and jet tagging~\cite{Cogan:2014oua} were written in 2014, only two years after AlexNet.
Since then there has been a remarkable growth in the number of papers applying deep learning to particle physics~\cite{hepmllivingreview}.
This thesis will provide a tour of some of these applications.
Concurrently, most other sub-fields of physics have taken interest in applying deep learning tools.
Cosmology, astrophysics, gravitational wave science, condensed matter physics, plasma physics, fluid dynamics, quantum information, and optical physics have all seen fruitful applications.
There are few areas of active physics research that have not been effected by the AlexNet moment in some sense.
While these applications have yet to produce a landmark discovery or new capability like the AlphaFold protein structure prediction model that was awarded the Nobel prize in chemisty in 2024, the breadth of these applications is a testatment to the power of deep learning for physics research.
% \footnote{Despite this active interest deep learning has yet to produce significant breakthroughs in other sub-fields of physics as well as particle physics. In contrast, deep learning tools have already produced a Nobel Prize in chemistry, for the AlphaFold protein structure prediction model in 2024.}.

In the remainder of the Introduction, I will outline the open problems in particle physics, and very briefly summarize the last 14 years of progress in machine learning and artificial intelligence.
This review serves two purposes.
The first is to give context to the very specific research detailed in this thesis.
The second is to reflect on the fact that all open problems in particle physics are long-standing, and that the list is essentially unchanged since the 2012 discovery of the Higgs.
These problems are extremely subtle and difficult, and many years of effort by very resourceful and intelligent scientists has so far proven insufficient to solve them.
By contrast, the field of machine learning and artificial intelligence is completely unrecognizable to itself in 2012.
The hope of this thesis, is that some of this dizzying progress can finally break the current impasse facing particle physics.

% ---------------------------------------------------------------
% Section: Open Problems in Particle Physics
% ---------------------------------------------------------------
%
% After completing the Standard Model with the Higgs, the field
% faces several deep, stubborn puzzles:
%   - Hierarchy / naturalness problem
%   - Dark matter (direct detection null results)
%   - Matter–antimatter asymmetry
%   - Neutrino masses and mixing
%   - Strong CP problem
%
% Key point: these problems have proved resistant to decades of
% targeted searches.  The LHC run-2 dataset is enormous (~140/fb)
% and growing, but extracting its full physics potential requires
% tools far beyond cut-based analyses.  This is where AI enters.

\section{Open problems in particle physics}
\label{sec:open_problems}

The most pressing open problems in particle physics can be understood as the need for a better understanding of Beyond the Standard Model (BSM) physics.
BSM physics is known to exist through multiple experimental observations that can not be explained by the particle content and forces contained in the Standard Model.
However the correct quantum field theory (or other type of theory consistent with the existing Standard Model) that explains these observations is unknown.
The most important observations are:

\begin{enumerate}
    \item{
        \textbf{Neutrino oscillations}: Neutrinos are massless under the Standard Model. However many iterations of neutrino experiments have verified that these particles ``oscillate'' between the three neutrino flavors as they propagate through space. This phenomenon is easily explained if the neutrinos carry a mass that is small enough to remain currently undetected by experiments, but the mechanism that produces this mass is unknown. This fascinating topic is not covered at all in this thesis, but current and next generation neutrino experiments are making active use of deep learning.
    }
    \item{
        \textbf{Dark matter}: A wide range of astrophysical and cosmological observations---galaxy rotation curves~\cite{Rubin:1970zza}, gravitational lensing, and the large-scale structure of the cosmic microwave background---indicate that roughly 27\% of the energy budget of the universe is composed of a non-luminous, non-baryonic form of matter.
        The Standard Model contains no candidate particle that could account for this ``dark matter''.
        Weakly Interacting Massive Particles (WIMPs) remain the most theoretically motivated candidate, but decades of direct detection experiments have returned null results, strongly constraining the simplest models.
        The LHC continues to search for dark matter production in missing-transverse-energy signatures, and deep learning has become an essential tool for both model-specific searches and model-agnostic anomaly detection approaches.
    }
    \item{
        \textbf{Matter--antimatter asymmetry}: The Standard Model predicts that the Big Bang should have produced equal amounts of matter and antimatter, yet the observable universe is composed overwhelmingly of matter.
        The known sources of CP violation in the Standard Model---the complex phase in the CKM quark mixing matrix and, to a lesser extent, the strong CP phase discussed below---are far too small to account for the observed asymmetry by many orders of magnitude~\cite{Sakharov:1967dj}.
        A new source of CP violation beyond the Standard Model is therefore required, and identifying it is one of the central motivations for precision flavor physics experiments such as LHCb and Belle~II.
    }
    \item{
        \textbf{Gravity}: General relativity provides an extraordinarily accurate classical description of gravity, yet all attempts to construct a consistent quantum field theory of gravity have so far failed.
        At the energy scales probed by the LHC ($\mathcal{O}$(TeV)) gravitational effects are negligible, so this is not a direct experimental puzzle for collider physics, but it represents the deepest incompleteness of our theoretical framework.
        Any truly unified description of nature must reconcile quantum mechanics and general relativity.
    }
    \item{
        \textbf{Hubble tension}: Two independent classes of measurement---the cosmic microwave background~\cite{Planck:2018vyg} and local distance-ladder observations~\cite{Riess:2021jrx}---yield values of the Hubble constant $H_0$ that disagree at the 4--5$\sigma$ level.
        Whether this tension reflects unaccounted systematic uncertainties or genuinely new physics beyond $\Lambda$CDM remains unresolved.
        If it is the latter, it would represent one of the clearest signals of beyond-Standard-Model physics in the cosmological sector.
    }
    \item{
        \textbf{Origin of the BAO scale}: Baryon Acoustic Oscillations (BAO)---the preferred clustering scale of galaxies imprinted by sound waves in the early universe---provide a powerful cosmological standard ruler.
        While the BAO scale is well measured and its physics within $\Lambda$CDM is well understood, tensions are beginning to emerge in BAO measurements from next-generation surveys such as DESI~\cite{DESI:2024mwx}.
        Understanding whether these tensions point to new physics in dark energy or the early universe is an active area of investigation.
    }
\end{enumerate}

The theoretical open problems are in some sense more troubling, because they are not observations that contradict the Standard Model but rather deep aesthetic and structural puzzles that suggest the theory is incomplete in ways that experiments have not yet confirmed.

\begin{enumerate}
    \item{
        \textbf{The hierarchy problem}: The Higgs boson mass receives quantum corrections from every particle it couples to.
        In the absence of fine-tuning or a symmetry protecting the Higgs mass, these corrections naturally drive the Higgs mass to the Planck scale ($\sim 10^{18}$~GeV), roughly 16 orders of magnitude above the observed value of 125~GeV.
        Supersymmetry, compositeness, and extra dimensions are the most studied solutions, but none has been confirmed experimentally.
        The absence of any BSM signal at the LHC has sharpened this ``naturalness'' tension considerably.
    }
    \item{
        \textbf{The strong CP problem}: Quantum chromodynamics (QCD) admits a CP-violating term in the Lagrangian parameterized by $\bar{\theta}$.
        Measurements of the neutron electric dipole moment constrain $|\bar{\theta}| \lesssim 10^{-10}$, yet there is no symmetry in the Standard Model that forces this parameter to be small.
        The most elegant proposed solution---the Peccei--Quinn mechanism, which predicts a new pseudo-Goldstone boson called the axion~\cite{Peccei:1977hh,Weinberg:1977ma}---also happens to be a compelling dark matter candidate, and axion searches are a very active experimental program.
    }
    \item{
        \textbf{The cosmological constant problem}: The observed energy density of the vacuum (dark energy) is roughly $10^{-47}$~GeV$^4$, yet naive estimates from quantum field theory predict a value $\sim 10^{120}$ times larger.
        This is arguably the worst fine-tuning problem in all of physics, and no compelling dynamical mechanism has been found to explain the observed small but non-zero value of the cosmological constant.
    }
    \item{
        \textbf{Accidental symmetries}: Several conservation laws of the Standard Model---baryon number $B$, and the individual lepton numbers $L_e$, $L_\mu$, $L_\tau$---are not imposed by the gauge structure of the theory but instead arise as ``accidental'' consequences of the field content and renormalizability.
        There is no deep theoretical reason these symmetries must be exact, and many well-motivated BSM theories predict their violation at some level.
        The experimental search for baryon-number violation (proton decay) and charged lepton flavor violation are active programs that would be unambiguous signs of new physics.
    }
    \item{
        \textbf{Scale of Yukawa couplings and the PMNS matrix}: The fermion masses in the Standard Model span nearly 12 orders of magnitude---from the electron at 0.511~MeV to the top quark at 173~GeV---yet the Higgs mechanism provides no explanation for this hierarchy of Yukawa couplings.
        Similarly, the neutrino mixing angles encoded in the PMNS matrix are large and apparently anarchic, in sharp contrast to the small mixing angles of the quark sector described by the CKM matrix.
        Why these parameters take the values they do, and whether there is an underlying flavor symmetry that explains them, remains unknown.
    }
    \item{
        \textbf{V-A structure of the weak interaction}: The weak interaction couples exclusively to left-handed fermions, a structure encoded in the $V-A$ (vector minus axial-vector) form of the weak current.
        While this is experimentally well established, there is no deep theoretical explanation within the Standard Model for why the weak force singles out left-handedness.
        Left-right symmetric extensions of the Standard Model restore parity at high energies, predicting new heavy gauge bosons, but none have been observed.
    }
    \item{
        \textbf{Calculability of QCD}: Quantum chromodynamics is in principle a complete theory of the strong force, but asymptotic freedom means it is only perturbatively tractable at high energies.
        At low energies---the regime of hadron masses, the proton structure, and non-perturbative QCD dynamics---the coupling becomes large and perturbation theory breaks down entirely.
        Lattice QCD provides a systematic non-perturbative approach but is computationally expensive and limited in scope.
        This difficulty is not merely academic: the ability to perform precise QCD calculations is a central bottleneck for extracting new physics signals from LHC data, and it is one of the most important problems that machine learning methods for unfolding and fast simulation aim to address.
    }
\end{enumerate}


% ---------------------------------------------------------------
% Section: The Machine Learning Revolution
% ---------------------------------------------------------------
%
% Contrast the state of ML in 2012 (AlexNet, ~60M parameters,
% trained on two GPUs) with today (LLMs at 10^{11} parameters,
% diffusion models generating photorealistic images).  The field
% is unrecognizable in 16 years.
%
% How this maps to HEP:
%   - Jet tagging: from cut-based to BDT to deep sets / GNN
%   - Anomaly detection: model-agnostic searches
%   - Fast simulation: GANs → flows → diffusion
%   - Unfolding: binned → machine-learned density ratios
%
% Key claim: the LHC is nearly an ideal laboratory for ML methods
% because (a) the data are high-dimensional and highly structured,
% and (b) we have extremely accurate synthetic data from MC
% simulation, enabling powerful supervised and semi-supervised
% approaches.

\section{The deep learning revolution and artificial intelligence}
\label{sec:ml_intro}

% ---------------------------------------------------------------
% Section: Thesis Roadmap
% ---------------------------------------------------------------
%
% Brief paragraph-per-chapter overview:
%   Ch.2 — QCD & jet substructure background
%   Ch.3 — ML methods (architectures through diffusion + density ratio)
%   Ch.4 — Jet tagging: ATLAS top-tagger paper + future outlook
%   Ch.5 — Unfolding: IBU → VLD → Omnifold
%   Ch.6 — Z+jets measurement with Omnifold
%   Ch.7 — Conclusion: ML as the future of collider physics

\section*{} % placeholder — remove once prose is written
